Об оценках на диаметр плоских множеств с целочисленными расстояниями полуобщего положения \footnote{Исследование выполнено за счет гранта Российского научного фонда (проект 19-11-00197).}

### 2021-10-31 semi-general-suplinear-bound/CHEBSB_Avdeev_rus.tex
Translate the intro

An \textit{integral point set} in a plane is a point set $M$ such that all the usual (Euclidean) distances between the
points of $M$ are integers and $M$ is not situated on a straight line.
Every integral point set consists of a finite number of points~\cite{anning1945integral,erdos1945integral};
thus, we denote the set of all planar integral point sets of $n$ points by
$\mathfrak{M}(2,n)$ (using the notation in~\cite{our-vmmsh-2018})
and define the diameter of $M\in\mathfrak{M}(2,n)$ in the following natural way:
\begin{equation}
	\operatorname{diam} M = \max_{A,B\in M} |AB|
	,
\end{equation}
where $|AB|$ denotes the Euclidean distance.
The symbol $\# M$ will be used for cardinality of $M$, that is the number of points in $M$ in our case.

Since every integral point set can obviously be dilated to a set of larger diameter,
minimal possible diameters of sets of given cardinality are of interest.
To be precise,
the following function was introduced in~\cite{kurz2008bounds,kurz2008minimum}:
\begin{equation}
	d(2,n) = \min_{M\in\mathfrak{M}(2,n)} \operatorname{diam} M
	.
\end{equation}

It turned out to be very easy to construct a planar integral point set of $n$ points with $n-1$ collinear ones and one point out of the line
(so-called \textit{facher} sets);
the same holds for 2 points out of the line (we refer the reader to~\cite{antonov2008maximal}, where some of such sets are called \textit{crabs})
and even for 4 points out of the line~\cite{huff1948diophantine}.
For $9\leq n\leq 122$, the minimal possible diameter is attained at a facher set~\cite{kurz2008bounds}.

\begin{Definition}
	A set $M\in\mathfrak{M}(2,n)$ is called to be in \textit{semi-general position},
	if no three points of $M$ are collinear.
	The set of all planar integral point sets in semi-general position
	is denoted by $\overline{\mathfrak{M}}(2,n)$.
\end{Definition}

Furthermore, the constructions of integral point sets in semi-general position of arbitrary cardinality
appeared in~\cite{harborth1993upper};
such sets are situated on a circle.
Also, there is a sophisticated construction of a circular integral point set of arbitrary cardinality
that gives the possible numbers of odd integral distances
between points in the plane~\cite{piepmeyer1996maximum}.

\begin{Definition}
	A set $M\in\overline{\mathfrak{M}}(2,n)$ is said to be a set in \textit{general position},
	if no four points of $M$ are concyclic.
	The set of all planar integral point sets in general position
	is denoted by $\dot{\mathfrak{M}}(2,n)$.
\end{Definition}

It remains unknown whether there are integral points sets in general position of arbitrary cardinality;
however, some sets $M\in \dot{\mathfrak{M}}(2,7)$ are known~\cite{kreisel2008heptagon,kurz2013constructing}.

The inequality
\begin{equation*}
	d(2,n) \leq \overline{d}(2,n) \leq \dot{d}(2,n)
	,
\end{equation*}
where
$
	\overline{d}(2,n) = \min_{M\in\overline{\mathfrak{M}}(2,n)} \operatorname{diam} M
$
and
$
	\dot{d}(2,n) = \min_{M\in\dot{\mathfrak{M}}(2,n)} \operatorname{diam} M
$,
is obvious; however, a more interesting relation holds:
\begin{equation*}
	c_1 n \leq d(2,n) \leq \overline{d}(2,n) \leq n^{c_2 \log \log n}
	.
\end{equation*}
The upper bound is established in~\cite{harborth1993upper}.
The lower bound firstly appeared in~\cite{solymosi2003note};
the largest known value for $c_1$ is $5/11$ for $n\geq 4$~\cite{my-pps-linear-bound-2019}.

There are some bounds for minimal diameter of planar integral point sets in some special positions.
Assuming that a planar integral point set contains many collinear points,
the following result holds.
\begin{Theorem}~\cite[Theorem 4]{kurz2008minimum}
	For $\delta > 0$, $\varepsilon > 0$, and $P\in\mathfrak{M}(2,n)$ with
	at least $n^\delta$ collinear points, there exists a $n_0 (\varepsilon)$
	such that for all $n \geq n_0 (\varepsilon)$ we have
	\begin{equation}
		\operatorname{diam} P \geq n^{\frac{\delta}{4 \log 2(1+\varepsilon)}\log \log n}
		.
	\end{equation}
\end{Theorem}
For diameter bounds for circular sets, we refer the reader to~\cite{bat2018number}.

Particular cases of planar integral point sets are also discussed
in~\cite[\S 5.11]{brass2006research},~\cite[\S D20]{guy2013unsolved},~\cite{our-pmm-2018},~\cite{our-ped-2018}.
For generalization to higher dimensions and the appropriate bounds, see~\cite{kurz2005characteristic,nozaki2013lower}.

In the present paper we give a special bound for planar integral point sets in semi-general position.
The condition of semi-general position is crucial in the given proof.

>>>>

\textit{Множеством с целочисленными расстояниями (МЦР)} на плоскости
называется такое множество точек $M$, что расстояние (евклидово) между любыми двумя точками из $M$
есть целое число, и при этом $M$ не является подмножеством никакой прямой.
Каждое МЦР состоит лишьиз конечного числа точек~\cite{anning1945integral,erdos1945integral};
поэтому мы можем обозначить множество всех множеств из $n$ точек с целочисленными расстояниями через
$\mathfrak{M}(2,n)$ (придерживаясь обозначений~\cite{our-vmmsh-2018} и в целом для краткости изложения)
и естественным образом определить диаметр множества $M\in\mathfrak{M}(2,n)$:
\begin{equation}
	\operatorname{diam} M = \max_{A,B\in M} |AB|
	,
\end{equation}
где $|AB|$ обозначает обычное (евклидово) расстояние между точками $A$ и $B$.
Через $\# M$ будем обозначать мощность множества $M$, которая в нашем случае всегда совпадает с количеством точек в $M$.

Поскольку любое МЦР можно очевидным образом растянуть
до множества большего диаметра,
интерес представляет в первую очередь минимальный возможный диаметр при заданной мощности множества.
Это рассуждение формализуется следующей функцией, введённой в~\cite{kurz2008bounds,kurz2008minimum}:
\begin{equation}
	d(2,n) = \min_{M\in\mathfrak{M}(2,n)} \operatorname{diam} M
	.
\end{equation}

Выяснилось, что достаточно просто сконструировать МЦР на плоскости,
все точки которого, кроме одной, лежат на некоторой прямой
(такие множества называются \textit{веерными});
аналогично --- для 2 точек вне прямой (см., напр., ~\cite{antonov2008maximal}, где такие множества названы \textit{крабовыми})
и даже для 4 точек вне прямой~\cite{huff1948diophantine}.
Для $9\leq n\leq 122$ минимальный возможный диаметр МЦР мощности $n$ достигаетя именно на веерном множестве~\cite{kurz2008bounds}.

\begin{Definition}
	Множество $M\in\mathfrak{M}(2,n)$ есть МЦР \textit{полуобщего положения},
	если никакие три точки из $M$ не лежат на одной прямой.
	Кратко будем писать, что $M\in \overline{\mathfrak{M}}(2,n)$.
\end{Definition}

В~\cite{harborth1993upper} приводится конструкция МЦР полуобщего положения для любой наперёд заданной мощности;
получаемое МЦР содержится в некоторой окружности.
Кроме того, в~\cite{piepmeyer1996maximum} приведена изящная конструкция МЦР $M$ любой наперёд заданной мощности,
содержащегося в окружности,
для любого возможного количества нечётных расстояний между точками $M$.

\begin{Definition}
	Множество $M\in\overline{\mathfrak{M}}(2,n)$ есть МЦР \textit{общего положения},
	если никакие четыре точки из $M$ не лежат на одной окружности.
	Кратко будем писать, что $M\in \dot{\mathfrak{M}}(2,n)$.
\end{Definition}

Остаётся неясным, любую ли мощность может имет МЦР общего положения;
однако, известны~\cite{kreisel2008heptagon,kurz2013constructing} примеры МЦР $M\in \dot{\mathfrak{M}}(2,7)$.

Очевидно неравенство
\begin{equation*}
	d(2,n) \leq \overline{d}(2,n) \leq \dot{d}(2,n)
	,
\end{equation*}
где
$
	\overline{d}(2,n) = \min_{M\in\overline{\mathfrak{M}}(2,n)} \operatorname{diam} M
$
и
$
	\dot{d}(2,n) = \min_{M\in\dot{\mathfrak{M}}(2,n)} \operatorname{diam} M
$.
Выполнено, однако, и более интересное соотношение:
\begin{equation*}
	c_1 n \leq d(2,n) \leq \overline{d}(2,n) \leq n^{c_2 \log \log n}
	.
\end{equation*}
Верхняя оценка установлена в~\cite{harborth1993upper}.
Нижняя оценка впервые была получена в~\cite{solymosi2003note};
наибольшее известное значение для $c_1$ равно $5/11$ (для $n\geq 4$)~\cite{my-pps-linear-bound-2019}.

Известны некоторые оценки на диаметр МЦР специального вида.
В частности, для множеств, в которых достаточно много точек лежит на одной прямой,
доказана следующая теорема.
\begin{Theorem}~\cite[Theorem 4]{kurz2008minimum}
	Для любых $\delta > 0$, $\varepsilon > 0$ и множества $P\in\mathfrak{M}(2,n)$,
	в котором не менее $n^\delta$ лежат на некоторой прямой, существует число $n_0 (\varepsilon)$
	такое, что для всех $n \geq n_0 (\varepsilon)$ верно неравенство
	\begin{equation}
		\operatorname{diam} P \geq n^{\frac{\delta}{4 \log 2(1+\varepsilon)}\log \log n}
		.
	\end{equation}
\end{Theorem}
Об оценках на диаметр МЦР,
содержащихся в окружности, можно прочесть в~\cite{bat2018number}.

Отдельные случаи МЦР на плоскости также обсуждались
в~\cite[\S 5.11]{brass2006research},~\cite[\S D20]{guy2013unsolved},~\cite{our-pmm-2018},~\cite{our-ped-2018}.
За обобщением на пространства более высоких размерностей и соответствующими оценками
отсылаем читателя к~\cite{kurz2005characteristic,nozaki2013lower}.

В настоящей статье мы доказываем отдельную оценку на диаметр множества с целочисленными расстояниями полуобщего положения.
Требование полуобщего положения является существенным для доказательства.
